
# OHBM Educational Course 2022 - Re-executing a publication

In this R notebook, you will be able to calculate case/control effect sizes
  and evaluate descriptive statistics for your cohort along the way.
  
  First, you need to make sure to set all the appropriate parameters. 
  
We will focus on a fixed regression here, but you will have some flexibility   
  with this script to test out other models
  
For this educational course we will run the following models:

+ your sample of 5 cases vs 5 controls from the ANT cohort (ds001907; Day et al 2022)
+ the full OpenNeuro ANT cohort (ds001907; Day et al 2022)
+ analysis of 3 other open cohorts used in the ENIGMA-PD publication (Laansma et al 2021)
+ analysis of your ANT cohort merged with the above for a total of 4 cohorts


### Read in the data and set the output folder and file names!
```{r}
datafile <- read.csv("all4sites_data_OHBM.csv"); #Read in the file
datafile <- read.csv("metr_SubCortical_OHBM.csv"); #Read in the file
datafile <- read.csv("OpenNeuro_results.csv"); #Read in the file
outFolder="./OpenNeuro_allROIs_woICV/" ; dir.create(outFolder,"/",showWarnings = FALSE);
cohortName="MyOHBMcohort-OpenNeuroFull-woICV"
eName="ENIGMA-PD"
```
```{r}
# Double check the column names of the file you read in:
columnnames = colnames(datafile);
colnames(datafile)
```

Do you need to calculate ICV? 
```{r}
   if (length(datafile[,which(columnnames=="ICV")]) == 0 
       && length(datafile[,which(columnnames=="csf..mm.3.")]) > 0 ) {
     print("It looks like you are working with the OHBM results and ICV is not included. We will add volumes of CSF, gray and white matter for you here -- this will serve as a proxy for ICV.")
     ICV=datafile$csf..mm.3. + datafile$white..mm.3. + datafile$gray..mm.3. 
     ncol=dim(datafile)[2] ;
     datafile[,ncol+1]=ICV ; colnames(datafile) <- c(columnnames,"ICV"); columnnames = colnames(datafile) } 
```

### This code is exclusively for case/control effect sizes as described by the Cohen's D statistic. It is important to ensure the code uses the proper column and indicator for cases vs controls: 
```{r}
# what is the column header for case/control or indicating diagnostic status?
caseColumnHeader= "diagnosis" #"Dx"
# how are cases coded (1,0,"patients","cases")?
caseIndicator=1 # in this code those not coded as cases are assumed to be controls
```

### We will be including age and sex in all models here. However, you can turn this feature off. 
```{r}
# what is the column header for age?
ageColumnHeader="age" # "Age" 
# what is the column header for sex?
sexColumnHeader="sex" # "Sex" 
# how are males coded (1,0,"M","males",-0.5)?
maleIndicator=1 #sex is included in the regression as a categorical variable, or factor
# would you like to include age, sex, and an age-x-sex interaction term in the regression?
includeAgeSexinModel=TRUE
# includeAgeSexInteractioninModel == TRUE # this option has been removed but you can add it back in (experts only -- be careful!)
```

### Do you have other covariates to include? We will ask about continuous and categorical variables separately
#### First, continuous:
```{r}
# how many continuous covariates other than age should be used from your spreadsheet?
otherCovariatesN=0
# list out the covariates and separate them with a semi-colon
otherCovariates="ICV" 
#note if your ROI is ICV, you will not want ICV as a covariate, so you'd have to run through the script again for the separate model
```
#### Next, categorical:
```{r}
# how many categorical covariates are in your spreadsheet other than sex?
## (examples might include site / scanner / study / different diagnostic groups) 
otherCatCovariatesN=0
# list out the categorical covariates (coded as factors) and separate them with a semi-colon
otherCatCovariates="site"
```

### What ROIs will you be looping over with the same statistical model? 
```{r}
ROINames<-c("Left-Accumbens-area (mm^3)","Left-Amygdala (mm^3)","Left-Caudate (mm^3)","Left-Hippocampus (mm^3)","Left-Pallidum (mm^3)","Left-Putamen (mm^3)","Left-Thalamus-Proper (mm^3)","Right-Accumbens-area (mm^3)","Right-Amygdala (mm^3)","Right-Caudate (mm^3)","Right-Hippocampus (mm^3)","Right-Pallidum (mm^3)","Right-Putamen (mm^3)","Right-Thalamus-Proper (mm^3)");

#ROINames<-c("Lthal","Rthal","Lcaud","Rcaud","Lput","Rput","Lpal","Rpal","Lhippo","Rhippo","Lamyg","Ramyg","Laccumb","Raccumb")

# ICV would require a different model since we will covary for ICV for the subcortical regions, so we'd run that separately 
#ROINames<-c("ICV")
ROINames<-make.names(ROINames)
#Get number of structures to test
Nrois=length(ROINames)
print(paste("We will be looping Cohen's D statistics over data from",Nrois,"regions of interest."))
```

### We will source some functions that we will use for Cohen's D calculations from a linear regression output. Feel free to take a look at these functions to see what they do
```{r}
# If this file is in the same folder as your current working directory, no need to change the input, otherwise please make sure the file with the appropriate path is sourced.
source('ENIGMA_cohensD_functions.R')
```

# Here we go! 

## Sanity checks:

```{r}
 # Check for duplicated SubjIDs that may cause issues with merging data sets.
 if(anyDuplicated(datafile[,1]) != 0) { stop('You have duplicate SubjIDs in your file or subject ID is not your first column.\nMake sure there are no repeat SubjIDs.') }

 # Check all ROIs are column headers in the input CSV
 missingROIS=NULL; l_missing=0;
 for (roi in 1:Nrois) {
   if (length(datafile[,which(columnnames==ROINames[roi])]) == 0) {
        print(ROINames[roi])
     missingROIS=paste(missingROIS,ROINames[roi])
     l_missing=l_missing+1; }
   }
 if (l_missing >= 1) { warning(paste("You are missing ROIs in your file.\n Please check.",sep=""))}
```
################################################################### 

<!-- ## Group statistics and plots -->

<!-- ```{r} -->
<!-- #install.packages('tidyverse','ggpubr','DT','RColorBrewer','scales') -->
<!-- library(ggpubr) -->
<!-- library('tidyverse') -->
<!-- library('DT') -->
<!-- library(scales) -->
<!-- library(RColorBrewer) -->

<!-- p1<- ggplot(data=datafile, aes(.data[[ageColumnHeader]], fill=factor(.data[[caseColumnHeader]])))  + -->
<!--   geom_histogram(aes(y = (..count..)/sum(..count..)),bins=10, position = "dodge",color="black"  ) +   -->
<!--   #geom_bar(aes(y = (..count..)/sum(..count..)),bins=20, position = "dodge",color="black" ) + -->
<!--         scale_y_continuous(labels = percent) + -->
<!--   #  theme(axis.text.x = element_text(angle = 90))+ -->
<!--    xlab("Age")+ ylab("Percent") -->

<!-- p2<- ggplot(data=datafile, aes(.data[[ageColumnHeader]], fill=factor(.data[[sexColumnHeader]]))) + -->
<!--    geom_histogram(aes(y = (..count..)/sum(..count..)),bins=10, position = "dodge",color="black" ) +   -->
<!--   #geom_bar(aes(y = (..count..)/sum(..count..)),bins=10, position = "stack",color="black" ) + -->
<!--         scale_y_continuous(labels = percent) + -->
<!--  #   theme(axis.text.x = element_text(angle = 90))+ -->
<!--    xlab("Age")+ ylab("Percent") -->


<!-- ggarrange(p1,p2, labels=c("Age by diagnosis", "Age by sex"),ncol=2,nrow=1,common.legend = FALSE, legend = "right") -->

<!-- ``` -->

################################################################### 
 
Here we initiate variables to save and print later
```{r}
#Store models for troubleshooting
models.all=NULL; # This will become a list where we store all of the models made by lm

#allocate empty vectors to store adjust effect sizes, se, ci, counts and pvalue
stats.d=matrix(NA,7,Nrois); 

featureNames=c("d","se","low.ci","up.ci","n.controls","n.cases","p-value")
colnames(stats.d) <- ROINames; rownames(stats.d) <- featureNames; 
```

Let's check the user inputs are consistent
```{r}
# check number indicated is the same as parsed text
covariates=parse(text=otherCovariates);
Ncov_parsed=length(covariates)
covariatesF=parse(text=otherCatCovariates)
Ncatcov_parsed=length(covariatesF)
```

```{r}
if (otherCovariatesN > 0) {
  if (Ncov_parsed != otherCovariatesN) { 
    stop(paste("You have indicated a different number of covariates than you have listed.\nPlease check listed variables are separated by a semicolon (;)",sep="")) } }
```

```{r}
if (otherCatCovariatesN > 0) {
  if (Ncatcov_parsed != otherCatCovariatesN) { 
    stop(paste("You have indicated a different number of categorical covariates than you have listed.\nPlease check listed variables are separated by a semicolon (;)",sep="")) } }
```

```{r}
#Loop through and perform each regression -- add covariates 
for (roi in 1:Nrois) {
  print(ROINames[roi])
  z=which(columnnames==ROINames[roi])
  rows2use=!is.na(datafile[,z])
  trait=datafile[rows2use,z]
  ## uncomment the following line to see how many samples are left in the ROI (after labeling NA for bad segmentations)
  #print(length(trait))
  if(length(trait)==0){
    next;}  # Skip the whole structure if there are no observations
  
  if (includeAgeSexinModel == TRUE ) {
# if (includeAgeSexInteractioninModel == TRUE ) {
# caseControl_covariatesforRFX = data.frame(NA,nrow=length(trait),ncol=otherCovariatesN+otherCatCovariatesN+4);
# ageXsex=(datafile[rows2use,sexColumnHeader] - mean(datafile[rows2use,sexColumnHeader])) *datafile[rows2use,ageColumnHeader]
#  caseControl_covariatesforRFX = datafile[rows2use,c(caseColumnHeader,ageColumnHeader,sexColumnHeader)]
#  origcolnames=colnames(caseControl_covariatesforRFX)
#  caseControl_covariatesforRFX[,4]=ageXsex
#  colnames(caseControl_covariatesforRFX)<-c(origcolnames,"ageXsex")
#  origcolnames=colnames(caseControl_covariatesforRFX)
#  string2run=paste(caseColumnHeader," + ",ageColumnHeader," + factor(",sexColumnHeader,") + ageXsex",sep="") }
  caseControl_covariatesforRFX = data.frame(NA,nrow=length(trait),ncol=otherCovariatesN+otherCatCovariatesN+3);
    caseControl_covariatesforRFX = datafile[rows2use,c(caseColumnHeader,ageColumnHeader,sexColumnHeader)]
  origcolnames=colnames(caseControl_covariatesforRFX)
    string2run=paste(caseColumnHeader," + ",ageColumnHeader," + factor(",sexColumnHeader,")",sep="") }
    
  else  {
  caseControl_covariatesforRFX = data.frame(NA,nrow=length(trait),ncol=otherCovariatesN+otherCatCovariatesN+1);
  caseControl_covariatesforRFX = datafile[rows2use,c(caseColumnHeader)]
  origcolnames=colnames(caseControl_covariatesforRFX)
  string2run=paste(caseColumnHeader) }

  if (otherCovariatesN > 0) {
    for (covariate in 1:otherCovariatesN) {
      covName=as.character(covariates[covariate])
      caseControl_covariatesforRFX[,length(caseControl_covariatesforRFX)+1] = datafile[rows2use,covName];
      colnames(caseControl_covariatesforRFX) = c(origcolnames,covName);
      origcolnames=colnames(caseControl_covariatesforRFX)
      string2run=paste(string2run,"+",covName,sep=" ") } }
  
  if (otherCatCovariatesN > 0) {
    for (covariate in 1:otherCatCovariatesN) {
      covName=as.character(covariatesF[covariate])
      caseControl_covariatesforRFX[,length(caseControl_covariatesforRFX)+1] = datafile[rows2use,covName];
      colnames(caseControl_covariatesforRFX) = c(origcolnames,covName);
      origcolnames=colnames(caseControl_covariatesforRFX)
      string2run=paste(string2run," + factor(",covName,") ",sep="") } }

## find and get rid of all rows with NAs in them 
    iNA=which(apply(caseControl_covariatesforRFX,1,function(x)any(is.na(x))));
    if (length(iNA) != 0 ) { caseControl_covariatesforRFX<-caseControl_covariatesforRFX[-which(apply(caseControl_covariatesforRFX,1,function(x)any(is.na(x)))),] }

#Run the model
  #attach main file
  attach(caseControl_covariatesforRFX,warn.conflicts = FALSE)
  eval(parse(text=paste("tmp=lm(trait ~ ", string2run, ")", sep='')))
  models.all[[roi]]=tmp #Store the model fit for future reference
  
  #subjects can be dropped if they are missing so we can get the precise number of controls/patients for each region tested
  stats.d[5,roi] = length(which(tmp$model[,2] != caseIndicator))
  stats.d[6,roi] = length(which(tmp$model[,2] == caseIndicator))
  
  #Convert the lm model to a summary format so we can extract statistics
  tmp=summary(tmp)
  ## uncomment the following line to see the full summary outputs of the models run
  #print(tmp)
  tstat=tmp$coefficients[2,3] # Get t-statistic from regression to convert to Cohens d
  pval=tmp$coefficients[2,4] # Get p-value from regression
  tstat.df=tmp$df[2]
  stats.d[7,roi] = pval
  
  #collect effect size data
  stats.d[1,roi]=partial.d(tstat,tstat.df, stats.d[5,roi], stats.d[6,roi])
  stats.d[2,roi]=se.d2(stats.d[1,roi],stats.d[5,roi], stats.d[6,roi])
  bound.cort=CI_95(stats.d[1,roi],stats.d[2,roi])
  stats.d[3,roi]=bound.cort[1]
  stats.d[4,roi]=bound.cort[2]
  rm(caseControl_covariatesforRFX)
}

```
```{r}
#save(stats.d, file=paste(outFolder,"/",eName, "_",cohortName,"_EffectSizes.Rdata", sep=""))
save(models.all, file=paste(outFolder,"/",eName, "_",cohortName,"_models.rda",sep=""))
write.csv(stats.d,paste(outFolder,"/",eName, "_",cohortName,"_EffectSizes.csv",sep=""))
stats.d

```
